### 2012-09-04 Dail_Mad_ext.tex
formatting

\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}

@@

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model---the
model for spatial and temporal variation in abundance. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  \label{eq:2}
  n_{it} \sim \mathrm{Bin}(N_{it}, p)
\end{equation}

@@

Differences in observers' ability to see, hear, or identify
birds has long been recognized as a potential source of error
in avian point count surveys such as the BBS (Robbins et
al. 1986, Sauer et al. 1994, Diefenbach et al. 2003, Alldredge
et al. 2007, Campbell and Francis 2011).  Estimating a
separate detection probability for each observer can be
difficult and reduces one's ability to estimate the quantities
of interest.  This problem is compounded by the fact that
observers differ greatly in the number of surveys they have
run (so that many observers' separate detection probabilities
could not be accurately estimated).

Current BBS trend estimators deal with this problem by
treating observer identity as a random (as opposed to a fixed)
effect (Link and Sauer 2002, Sauer and Link 2011).  This
allows observer-specific differences in detection probability,
but assumes that observers are selected at random from a pool
of potential observers.  Models that contain both random and
fixed effects are referred to as mixed models.  Often in mixed
models (as in this case), the random effect is modeled not
because it of interest in itself, but to avoid bias in the
estimates of the fixed effects.

To include random observer effects in DM models and these extensions,
equation 5 can be modified to:
(15)
(16)
where nijt is the number of bird recorded at site i by observer j in
year t, pj is observer-specific detection probability, $\mu_p$ is the mean
detection probability (on the logit scale), and $\sigma_p$ is the standard
deviation of the random observer effects (also on the logit scale).
We have implemented random observer effects in the Bayesian framework
using program JAGS (version 3.2.0; Plummer 2003) with the R (R
Development Core Team 2011) package rjags (Plummer 2011) as an
interface.

@@

The zero-inflated Poisson distribution can be applied to not only
initial abundance but also to recruitment and population growth terms.
For example, the recruitment term of the constant DM model (equation
4) can be modified as follows:
(17)
When both initial abundance and dynamics are modeled as zero-inflated,
$\psi$ can be modeled as time-varying or not (in either case, $\psi$ can also be
modeled as varying over space).  In the latter case, it makes sense
for the model to "remember" which sites are inside the range of the
species between the initial abundance stage and each time step of the
dynamics stage (see code sample?).  Adding zero-inflation to the
dynamics of the autoregressive, trend, Ricker, or Gompertz models
(without immigration) should have no effect when initial abundance is
also zero-inflated unless $\psi$ is time-varying (because these dynamic
models cannot recover from zero abundance).  We have implemented
zero-inflated dynamics in the Bayesian framework using program JAGS
(version 3.2.0; Plummer 2003)

@@

We simulated data for 100 sites over 40 years.  All
simulations assumed initial abundance was Poisson distributed
and no covariates affected initial abundance, dynamics, or
detection probability.  Our first series of simulations
assumed dynamics were exponential.  We ran 500 simulations for
each combination of low, medium, and high $\Lambda$, $r$, and
$p$ (Table 1; 27 total combinations).

@@

We estimated the parameters for each simulation using the same
initial abundance (Poisson) and dynamics models as were
simulated, implemented in the unmarked library in R.  When run
in a maximum likelihood framework, these models require a
maximum abundance to integrate over (Royle 2004, Dail and
Madsen 2011); we used 150.  We report bias of estimates, root
mean squared error, and coverage (percentage of 95% confidence
% intervals for parameters that
% overlap the true values).
Ideas:
I.	Test importance of random effects modeling, and other differences between JAGS and unmarked
II.	Model with low p values, or a range of p values
III.	Ability to distinguish correct initial abundance and dynamics models
IV.	When can you detect DD if it exists (power)?
V.	Building on the last two: what if your model set doesn't
include the correct DD model?  When will the wrong DD model be favored
over the wrong DID model?

@@

We applied these models to North American Breeding Bird Survey (BBS)
data from 1966-2010 for two species in the bordering US states
Maryland and Virginia.  For our focal species, we selected ovenbirds
(Seiurus aurocapilla), an abundant and widespread forest breeding
migrant with a stable or increasing trend in the region (CITE) and
golden-winged warblers (Vermivora chrysoptera), which only breeds in
the western parts of these states and has been declining in the region
(CITE).  MORE INFORMATION ABOUT THE SPECIES?

GENERAL ABOUT BBS DATA.  We only used data marked as acceptable for
use in the annual BBS analysis.  We summed the total birds of each
species seen on each route and year and used the routes (rather than
the individual stops) as our sites.
Strong winds can interfere with point count observers' ability to hear
birds (Simons et al. 2007)(cite); we tested the effects of wind speed
on detection probability.  BBS volunteers record wind conditions at
the beginning and end of each route on the Beaufort Scale (start and
end wind; Robbins et al. 1986)(start and end wind; 0-9; cite).  When
start or end wind was not recorded we imputed those values with the
start or end mean.  We took the mean of start and end wind for each
route and put this average wind scale value into four categories: 0 $\leq$
wind $<$ 1; 1 $\geq$ wind $<$ 2; 2 $\leq$ wind $<$ 3; and wind $\geq$ 3 (maximum of 3.5).
Following Link and Sauer (2002), we also included the first time an
observer ran a route as a predictor variable for detection
probability.

@@

We ran the top ranked models from the maximum likelihood analyses in a
Bayesian framework.  We added random observer effects and, where
appropriate, zero-inflated dynamics.  WILL ADD MORE HEREWe used
non-informative priors.  We tested for lack of convergence using X
Markov chains for each model (Gelman and Rubin 1992).  For each chain
we sampled the MCMC for Y iterations, after Z tuning samples.  Latent
inclusion factors for temperature effects?

@@

>>>>

\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}

@@

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model---the
model for spatial and temporal variation in abundance. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  n_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}

@@

Differences in observers' ability to see, hear, or identify
birds has long been recognized as a potential source of error
in avian point count surveys such as the BBS
\citep{robbins_etal:1986,diefenbach_etal:2003,sauer_etal:1994auk,alldredge_etal:2007auk,campbell_francis:2011}.
Estimating a separate detection probability for each observer can be
difficult and reduces one's ability to estimate the quantities
of interest.  This problem is compounded by the fact that
observers differ greatly in the number of surveys they have
run (so that many observers' separate detection probabilities
could not be accurately estimated).

Current BBS trend estimators deal with this problem by
treating observer identity as a random (as opposed to a fixed)
effect \citep{link_sauer:2002,sauer_link:2011}.
%(Link and Sauer 2002, Sauer and Link 2011).
This allows observer-specific differences in detection probability,
but assumes that observers are selected at random from a pool
of potential observers.  Models that contain both random and
fixed effects are referred to as mixed models.  Often in mixed
models (as in this case), the random effect is modeled not
because it of interest in itself, but to avoid bias in the
estimates of the fixed effects.

To include random observer effects in DM models and these extensions,
Eq.~\ref{eq:p1} can be modified to:
\begin{gather}
n_{ijt} \sim \mathrm{Bin}(N_{it}, p_j) \nonumber \\
\mathrm{logit}(p_j) \sim \mathrm{Normal}(\mu_p, \sigma_p)
\label{eq:pobs}
\end{gather}
where $n_{ijt}$ is the number of bird recorded at site $i$ by observer $j$ in
year $t$, $p_j$ is observer-specific detection probability, $\mu_p$ is the mean
detection probability (on the logit scale), and $\sigma_p$ is the standard
deviation of the random observer effects (also on the logit scale).
We have implemented random observer effects in the Bayesian framework
using program \textbf{JAGS} \citep[version 3.2.0]{plummer:2003}
% (version 3.2.0; Plummer 2003)
with the \textbf{R} (R
Development Core Team 2011) package \texttt{rjags} \citep{plummer:2011}
%(Plummer 2011)
interface.

@@

The zero-inflated Poisson distribution can be applied to not only
initial abundance but also to recruitment and population growth terms.
For example, the recruitment term of the constant DM model (equation
4) can be modified as follows:
\begin{equation}
G_{it} \sim \left\{
\begin{aligned}
0 &\; \text{with probability} \; \psi \\
\mathrm{Poisson}(\gamma) &\; \text{with probability} \; (1-\psi)
\end{aligned} \right.
\label{eq:ZIPts}
\end{equation}
When both initial abundance and dynamics are modeled as zero-inflated,
$\psi$ can be modeled as time-varying or not (in either case, $\psi$ can also be
modeled as varying over space).  In the latter case, it makes sense
for the model to "remember" which sites are inside the range of the
species between the initial abundance stage and each time step of the
dynamics stage (see code sample?).  Adding zero-inflation to the
dynamics of the autoregressive, trend, Ricker, or Gompertz models
(without immigration) should have no effect when initial abundance is
also zero-inflated unless $\psi$ is time-varying (because these dynamic
models cannot recover from zero abundance).  We have implemented
zero-inflated dynamics in the Bayesian framework using program
\textrm{JAGS} \citep[version 3.2.0]{plummer:2003}.
%(version 3.2.0; Plummer 2003)

@@

We simulated data for 100 sites over 40 years.  All
simulations assumed initial abundance was Poisson distributed
and no covariates affected initial abundance, dynamics, or
detection probability.  Our first series of simulations
assumed dynamics were exponential.  We ran 1000 simulations for
each combination of low, medium, and high $\Lambda \in \{1,5,10\}$, $r
\in \{-0.005, 0, 0.005\}$, and
$p \in \{0.05, 0.25, 0.5\}$. % (Table 1; 27 total combinations).

@@

We estimated the parameters for each simulation using the same
initial abundance (Poisson) and dynamics models as were
simulated, implemented in the unmarked library in \textbf{R}.  When run
in a maximum likelihood framework, these models require a
maximum abundance to integrate over
\citep{royle:2004biom,dail_madsen:2011};
%(Royle 2004, Dail and Madsen 2011);
we used 200.  We report bias of estimates, root
mean squared error, and coverage (percentage of 95\% confidence
intervals for parameters that
overlap the true values).

Ideas:
I.	Test importance of random effects modeling, and other differences between JAGS and unmarked
II.	Model with low p values, or a range of p values
III.	Ability to distinguish correct initial abundance and dynamics models
IV.	When can you detect DD if it exists (power)?
V.	Building on the last two: what if your model set doesn't
include the correct DD model?  When will the wrong DD model be favored
over the wrong DID model?

@@

We applied these models to North American Breeding Bird Survey (BBS)
data from 1966-2010 for two species in the bordering US states
Maryland and Virginia.  For our focal species, we selected ovenbirds
(\textit{Seiurus aurocapilla}), an abundant and widespread forest breeding
migrant with a stable or increasing trend in the region (CITE) and
golden-winged warblers (\textit{Vermivora chrysoptera}), which only breeds in
the western parts of these states and has been declining in the region
(CITE).  MORE INFORMATION ABOUT THE SPECIES?

GENERAL ABOUT BBS DATA.  We only used data marked as acceptable for
use in the annual BBS analysis.  We summed the total birds of each
species seen on each route and year and used the routes (rather than
the individual stops) as our sites.
Strong winds can interfere with point count observers' ability to hear
birds (\citep{simons_etal:2007}); we tested the effects of wind speed
on detection probability.  BBS volunteers record wind conditions at
the beginning and end of each route on the Beaufort Scale
\citep[start and end wind 0-9]{robbins_etal:1986}.
%(start and end wind; Robbins et al. 1986)(start and end wind; 0-9; cite).
When
start or end wind was not recorded we imputed those values with the
start or end mean.  We took the mean of start and end wind for each
route and put this average wind scale value into four categories: 0 $\leq$
wind $<$ 1; 1 $\geq$ wind $<$ 2; 2 $\leq$ wind $<$ 3; and wind $\geq$ 3 (maximum of 3.5).
Following \citet{link_sauer:2002},
%Link and Sauer (2002),
we also included the first time an
observer ran a route as a predictor variable for detection
probability.

@@

We ran the top ranked models from the maximum likelihood analyses in a
Bayesian framework.  We added random observer effects and, where
appropriate, zero-inflated dynamics.  WILL ADD MORE HEREWe used
non-informative priors.  We tested for lack of convergence using X
Markov chains for each model \citep{gelman_rubin:1992}.
%(Gelman and Rubin 1992).
For each chain
we sampled the MCMC for Y iterations, after Z tuning samples.  Latent
inclusion factors for temperature effects?

@@

\begin{figure}
  \centering
  \includegraphics[height=8in]{figs/exp_hists}
  \caption{Histograms of 1000 parameter estimates for each of 27 simulation cases. The verticle lines are the data-generating values.}
\label{fig:exp_hists}
\end{figure}
